Qualitative Probabilistic Programming

### 2015-05-30 paper/paper.tex
paper

  Here is an example of a Quipp program to cluster 2d points into 3 clusters:

  \begin{verbatim}
var nclusters = 3;
var dim = 2;

@@

testParamInference(function(randFunction) {
  var pointType = Vector(dim, Double);
  var getPoint = randFunction(Categorical(nclusters), pointType);
  return function() {
    return repeat(n, function(i) {
      var cluster = randomInteger(nclusters);
      return observe('point' + i, getPoint, cluster);
    });
  };
});
  \end{verbatim}

  It is written as a generative model producing the observations.  Notice that we declared two types (\texttt{PointType} and \texttt{ClusterType})
  and one random function \texttt{getPoint}).
  Type annotations are necessary for random functions.  The type
  \texttt{Vector(2, Double)} expresses the fact that the points are 2-dimensional,
  and the type \texttt{Categorical(3)} expresses the fact that there are 3
  possible clusters (so a cluster may be either 0, 1, or 2).  We assume that
  the distribution over clusters is uniform, but we do not know the distribution
  of points in each cluster.
  We will fill in the
  \texttt{getPoint} function with a learned function that will take a random
  sample from the given cluster.

@@

{\small
\begin{verbatim}
nfeatures = 10
FeaturesType = Vector(nfeatures, Bool)

get_class = rand_function(Unit, Bool)

class_1_features = rand_function(Unit, FeaturesType)
class_2_features = rand_function(Unit, FeaturesType)

def sample():
  which_class = get_class()
  if which_class:
    features = class_1_features()
  else:
    features = class_2_features()
  return (which_class, features)
\end{verbatim}
}

@@

  \subsection{Factor analysis}
{\small
\begin{verbatim}
num_components = 3
point_dim = 5

ComponentsType = Vector(num_components, Double)
PointType = Vector(point_dim, Double)

get_point = rand_function(ComponentsType, PointType)

def sample():
  components = [normal(0, 1)
                for i in range(num_components)]
  return (comenents, get_point(components))
\end{verbatim}
}

@@

(figure should show accuracy by EM iteration.  Accuracy can be measured by distances between different factor coefficients, when ordered according to some permutation)
  \subsection{Hidden Markov model}
{\small
\begin{verbatim}
num_states = 5
num_observations = 3
sequence_length = 10

StateType = Categorical(num_states)
ObsType = Categorical(num_observations)

trans_fun = rand_function(StateType, StateType)
obs_fun = rand_function(StateType, ObsType)

def sample():
  states = [uniform(num_states)]
  for i in range(sequence_length - 1):
    states.append(trans_fun(states[-1]))
  return (states, [obs_fun(s) for s in states])
\end{verbatim}
}

In this example, we use the unknown function \texttt{trans\_fun} for state transitions and \texttt{obs\_fun} for observations.  This means that we will learn both the state transitions and the observation distribution.

\subsection{Latent Dirichlet allocation}
{\small
\begin{verbatim}
n_classes = 10
n_words = 100
n_words_per_document = 1000

ClassType = Categorical(n_classes)
WordType = Categorical(n_words)

class_to_word = rand_function(ClassType, WordType)

def sample():
    which_class = uniform_categorical(n_classes)
    words = [class_to_word(which_class)
             for i in range(n_words_per_document)]
    return (which_class, words)
\end{verbatim}
}

In this example, we use the unknown function \texttt{class\_to\_word} to map classes to word distributions.  Note that each column of the matrix of parameters for \texttt{class\_to\_word} will represent a categorical distribution over words, and there will be one column for each class.

@@

{\small
\begin{verbatim}
input_dim = 100
hidden_dim = 20
output_dim = 2

InputType = Categorical(input_dim)
HiddenType = Categorical(hidden_dim)
OutputType = Categorical(output_dim)

input_to_hidden = rand_function(InputType, HiddenType)
hidden_to_output = rand_function(HiddenType, OutputType)

def sample(input_layer):
  hidden_layer = input_to_hidden(input_layer)
  output_layer = hidden_to_output(hidden_layer)
  return ((), output_layer)
\end{verbatim}
}

>>>>

  Here is an example of a Quipp program to cluster 2d points into 3 clusters:
  \begin{verbatim}
var nclusters = 3;
var dim = 2;

@@

testParamInference(function(randFunction) {
  var pointType = Vector(dim, Double);
  var getPoint = randFunction(Categorical(nclusters), pointType);
  return function() {
    repeat(n, function(i) {
      var cluster = randomInteger(nclusters);
      observe('point' + i, getPoint, cluster);
    });
  };
});
  \end{verbatim}

  It is written as a generative model producing the observations.  Notice that
  we declared two types (\texttt{PointType} and \texttt{ClusterType}) and one
  random function \texttt{getPoint}).  Type annotations are necessary for random
  functions.  The type \texttt{Vector(2, Double)} expresses the fact that the
  points are 2-dimensional, and the type \texttt{Categorical(3)} expresses the
  fact that there are 3 possible clusters (so a cluster may be either 0, 1, or
  2).    We assume the distribution over clusters is uniform, but we do not know
  the distribution of points in each cluster.  We will fill in the
  \texttt{getPoint} function with a learned function that will take a random
  sample from the given cluster.  The \texttt{observe} function allows us to
  observe data; here it says that the observation named \texttt{point3} is equal
  to the result of calling \texttt{getPoint(cluster)}, where \texttt{cluster} is
  cluster number 3.

  To demonstrate, let us run this example on a dataset consisting of 150 points (TODO cite).  When we run the program on this data, we infer the parameters to the random function \texttt{getPoint}.
  In this case, \texttt{getPoint} is a linear function with Gaussian noise, so it will naturally
  split the data into 2 clusters with equal variance:
  \begin{center}
    \includegraphics[scale=0.5]{../plots/irisclusters_orig.png}
  \end{center}

  The first cluster is at (6.3, 5.0) and the second is at (5.0, 1.6).  They both
  have a standard deviation of 0.54 in the x direction and 0.69 in the y direction.
  We could use these parameters to fill in the generative model:
  \begin{verbatim}
    return repeat(n, function(i) {
      var cluster = randomInteger(nclusters);
      return [gaussian(cluster == 0 ? 6.3 : 5.0, 0.54),
              gaussian(cluster == 0 ? 5.0 : 1.6, 0.69)];
    });
  \end{verbatim}
  This model is estimated to assign probability density $e^{-393}$ to the data, or 0.0728 per point.

  Note that, because the two clusters are forced to have the same variance, they do not fit the data well,
  since the data has a different shape in each location.  To fix this problem, we can substitute the following
  model:
  \begin{verbatim}
testParamInference(function() {
  var pointType = Vector(dim, Double);
  var getPointFunctions = repeat(nclusters, function(i) {
    return randFunction(pointType);
  });
  return function() {
    return repeat(n, function(i) {
      var cluster = randomInteger(nclusters);
      return observe('point' + i, getPointFunctions[cluster]);
    });
  };
});
  \end{verbatim}
  Using this model, we get the following clusters:
  \begin{center}
    \includegraphics[scale=0.5]{../plots/irisclusters_indep.png}
  \end{center}
  This model (with the parameters filled in) is estimated to assign probability density $e^{-337}$ to the data, or 0.1058 per point, which means it
  fits the data better.

@@

{\small
\begin{verbatim}
var nfeatures = 10;
var featuresType = Vector(nfeatures, Bool);
var nclasses = 2;
var n = 50;

testParamInference(function(randFunction) {
  var class0Features = randFunction(featuresType);
  var class1Features = randFunction(featuresType);

  var getFeatures = randFunction(Categorical(nclusters), pointType);
  return function() {
    repeat(n, function(i) {
      var whichClass = randomInteger(2);
      if (whichClass == 0) {
        observe('features' + i, class0Features);
      } else {
        observe('features' + i, class1Features);
      }
    });
  };
});
\end{verbatim}
}

@@

  \subsection{Factor analysis}
{\small
\begin{verbatim}

var nfactors = 2;
var dim = 4;

var n = 50;

testParamInference(function(randFunction) {
  var factorType = Vector(nfactors, Double);
  var pointType = Vector(dim, Double);
  var getPoint = randFunction(factorType, pointType);
  var getFactors = function() {
    return repeat(nfactors, function() { return gaussian(0, 1); });
  };
  return function() {
    return repeat(n, function(i) {
      var factors = getFactors();
      return observe('point' + i, getPoint, factors);
    });
  };
});
\end{verbatim}
}

@@

  \subsection{Hidden Markov model}
{\small
\begin{verbatim}

var nstates = 2;
var nobs = 3;

var chainLength = 8;

var n = 30;

testParamInference(function(randFunction) {
  var stateType = Categorical(nstates);
  var obsType = Categorical(nobs);
  var transFun = randFunction(stateType, stateType);
  var obsFun = randFunction(stateType, obsType);

  var observeStates = function(sampIndex, startState, i) {
    if (i == chainLength) {
      return [];
    } else {
      observe('obs' + sampIndex + '_' + i, obsFun, startState);
      return [startState].concat(observeStates(sampIndex, transFun(startState), i+1));
    }
  };
  return function() {
    return repeat(n, function(sampIndex) {
      return observeStates(sampIndex, randomInteger(nstates), 0);
    });
  };
});
\end{verbatim}
}

In this example, we use the unknown function \texttt{transFun} for state transitions and \texttt{obsFun} for observations.  This means that we will learn both the state transitions and the observation distribution.

\subsection{Latent Dirichlet allocation}
{\small
\begin{verbatim}
var nClasses = 2;
var nWords = 100;
var maxWordsPerDocument = 1000;
var nDocuments = 20;

var classType = Categorical(nClasses);
var wordType = Categorical(nWords);

testParamInference(function(randFunction) {
  var classToWord = randFunction(ClassType, WordType);
  return function() {
    return repeat(nDocuments, function(docIndex) {
      var whichClass = randomInteger(nClasses);
      var nWords = observe('len' + docIndex, randomInteger, maxWordsPerDocument);
      repeat(nWordsPerDocument, function(wordIndex) {
        observe('word' + doc + '_' + wordIndex, classToWord, whichClass);
      });
      return whichClass;
    });
  };
});
\end{verbatim}
}

In this example, we use the unknown function \texttt{classToWord} to map classes to word distributions.  Note that each column of the matrix of parameters for \texttt{classToWord} will represent a categorical distribution over words, and there will be one column for each class.

@@

{\small
\begin{verbatim}
var inputDim = 100;
var hiddenDim = 20;
var outputDim = 2;

var inputType = Categorical(inputDim);
var hiddenType = Categorical(hiddenDim);
var outputType = Categorical(outputDim);

var inputs = [...];

testParamInference(function(randFunction) {
  var inputToHidden = randFunction(inputType, hiddenType);
  var hiddenToOutput = randFunction(hiddenType, outputType);

  return function() {
    return repeat(inputs.length, function(sampIndex) {
      var hiddenLayer = inputToHidden(inputs[sampIndex]);
      observe('output' + sampIndex, hiddenToOutput, hiddenLayer);
    });
  };
});
\end{verbatim}
}
